---
title: "Pear quality and longitudinal growth: complete exploratory analysis"
author: "Tanjim Hossain"
date: "2026-04-13"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
---

This portfolio exploratory workflow is retained below in full.
Paths and selected display labels have been adapted for this repository.
The field `quality` in the size dataset means **pear size in cm**, not the harvest quality index.
Exploratory error bars describe the observations; they are not cluster-adjusted model confidence intervals.
The complete-case section is a sensitivity dataset, not a claim that dropout is ignorable.

## Packages

```{r}
library(tidyverse)
library(ggplot2)
library(lattice)
library(viridis)
```


## Harvest data: structure and missingness

```{r}
main.data <- read.csv("data/raw/G12.outcome.data.csv",
                      sep = ",")
str(main.data)

sapply(main.data, function(x) sum(is.na(x)))
```

## Quality distributions across climate, species and location

```{r}
ggplot(main.data, aes(climate, quality_idx, colour = location))+
  geom_boxplot()+
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  labs(y="quality index")+
  theme_bw()

ggplot(main.data, aes(species, quality_idx, colour = location))+
  geom_boxplot()+
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  labs(y="quality index")+
  theme_bw()

ggplot(main.data, aes(location, quality_idx))+
  geom_boxplot()+
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  labs(y="quality index")+
  theme_bw()

ggplot(main.data, aes(climate, quality_idx, colour = species))+
  geom_boxplot()+
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  labs(y="quality index")+
  theme_bw()

mean_data <- main.data %>% 
  group_by(location) %>% 
  summarise(mean_quality = mean(quality_idx))

ggplot(main.data, aes(climate, quality_idx, colour = species)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75))+
  geom_hline(data = mean_data,
             aes(yintercept = mean_quality),
             linetype = "dashed",
             color = "black") +
  facet_wrap(~location) +
  theme_bw()
```

## Exploratory interactions for continuous quality

```{r}
interaction.plot(x.factor = main.data$species, 
                 trace.factor = main.data$climate, 
                 response = main.data$quality_idx,
                 trace.label = "climate",
                 type = "b",
                 xlab = "species", 
                 ylab = "quality index",
                 ylim = c(45, 65))

interaction.plot(x.factor = main.data$species, 
                 trace.factor = main.data$location, 
                 response = main.data$quality_idx,
                 trace.label = "location",
                 type = "b",
                 xlab = "species", 
                 ylab = "quality index",
                 ylim = c(45, 65))

interaction.plot(x.factor = main.data$location, 
                 trace.factor = main.data$climate, 
                 response = main.data$quality_idx,
                 trace.label = "climate",
                 type = "b",
                 xlab = "location", 
                 ylab = "quality index",
                 ylim = c(45, 65))

ggplot(main.data, aes(x = climate, y = quality_idx, color = species, group = species)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_grid(~location) + 
  theme_minimal() +
  labs(y = "Avg Quality Index")
```

## Binary endpoint: quality index at least 55

```{r}
main.data2 <- main.data %>% 
  mutate(quality_b = if_else(quality_idx >= 55,1,0))

# write.csv(main.data2,"data/processed/quality_binary.csv",
#           row.names = FALSE)
```

## Tree-level proportions and subgroup comparisons

```{r}
main.data2 %>%
  group_by(location, tree_id, species, climate) %>%
  summarise(prob_success = mean(quality_b), .groups = "drop") %>% 
  ggplot(aes(location, prob_success, fill = species))+
  geom_boxplot()+
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75))+
  facet_wrap(~climate)+
  labs(y = "probability of success") +
  theme_bw()

main.data2 %>%
  count(location, species, climate, quality_b) %>%
  group_by(location, species, climate) %>%
  mutate(percent = n / sum(n)) %>% 
  ggplot(aes(climate, percent, fill = as.factor(quality_b))) +
  geom_bar(stat = "identity",position = "dodge") +
  facet_wrap(~location) +
  scale_fill_manual(values = c("gray70", "skyblue"), labels = c("Bad", "Good")) +
  labs(y = "proportion of quality", fill = "Outcome") +
  theme_minimal()

main.data2 %>%
  count(location, species, climate, quality_b) %>%
  group_by(location, species, climate) %>%
  mutate(percent = n / sum(n)) %>% 
  ggplot(aes(climate, percent, fill = as.factor(quality_b))) +
  geom_bar(stat = "identity",position = "dodge") +
  facet_wrap(~species) +
  scale_fill_manual(values = c("gray70", "skyblue"), labels = c("Bad", "Good")) +
  labs(y = "proportion of quality", fill = "Outcome") +
  theme_minimal()

main.data2 %>%
  count(location, species, climate, quality_b) %>%
  group_by(location, species, climate) %>%
  mutate(percent = n / sum(n)) %>% 
  ggplot(aes(climate, percent, fill = as.factor(quality_b))) +
  geom_bar(stat = "identity",position = "dodge") +
  facet_grid(location~species) +
  scale_fill_manual(values = c("gray70", "skyblue"), labels = c("Bad", "Good")) +
  labs(y = "proportion of quality", fill = "Outcome") +
  theme_minimal()
```

```{r}
main.data2 %>%
  group_by(location, tree_id, species, climate) %>%
  summarise(prob_success = mean(quality_b), .groups = "drop") %>% 
  ggplot(aes(climate, prob_success, fill = location)) +
  geom_boxplot() + 
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  labs(y="probability of success") +
  theme_bw()

main.data2 %>%
  group_by(location, tree_id, species, climate) %>%
  summarise(prob_success = mean(quality_b), .groups = "drop") %>% 
  ggplot(aes(climate, prob_success, fill = species)) +
  geom_boxplot() + 
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  labs(y="probability of success") +
  theme_bw()

main.data2 %>%
  group_by(location, tree_id, species, climate) %>%
  summarise(prob_success = mean(quality_b), .groups = "drop") %>% 
  ggplot(aes(species, prob_success, fill = location)) +
  geom_boxplot() + 
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  labs(y="probability of success") +
  theme_bw()

main.data2 %>%
  group_by(location, tree_id, species, climate) %>%
  summarise(prob_success = mean(quality_b), .groups = "drop") %>% 
  ggplot(aes(climate, prob_success, fill = species)) +
  geom_boxplot() + 
  stat_summary(fun = mean, geom = "point",
               position = position_dodge(width = 0.75)) +
  facet_wrap(~location) +
  labs(y="probability of success") +
  theme_bw()

main.data2 %>%
  group_by(location, tree_id, species, climate) %>%
  summarise(prob_success = mean(quality_b), .groups = "drop") %>% 
  ggplot(aes(x = location, y = prob_success, fill = species)) +
  geom_boxplot(position = position_dodge(width = 0.8)) + 
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black",
               position = position_dodge(width = 0.8)) +
  facet_grid(species ~ climate) +
  labs(y = "probability of success") +
  theme_bw()
```

## Exploratory interactions for binary quality

```{r}
interaction.plot(x.factor = main.data2$species, 
                 trace.factor = main.data2$climate, 
                 response = main.data2$quality_b,
                 trace.label = "climate",
                 type = "b",
                 xlab = "species", 
                 ylab = "probability of success",
                 ylim = c(0.30, 0.7))

interaction.plot(x.factor = main.data2$species, 
                 trace.factor = main.data2$location, 
                 response = main.data2$quality_b,
                 trace.label = "location",
                 type = "b",
                 xlab = "species", 
                 ylab = "probability of success",
                 ylim = c(0.30, 0.7))

interaction.plot(x.factor = main.data2$location, 
                 trace.factor = main.data2$climate, 
                 response = main.data2$quality_b,
                 trace.label = "climate",
                 type = "b",
                 xlab = "location", 
                 ylab = "probability of success",
                 ylim = c(0.30, 0.7))

```



## Longitudinal source and missing measurements

```{r}
time.data <- read.csv("data/raw/G12.size.data.csv",
                      sep = ",")
str(time.data)

sapply(time.data, function(x) sum(is.na(x)))
time.data[!complete.cases(time.data), ]
library(naniar)
miss_var_summary(time.data)
```

Missing measurements follow a monotone dropout pattern. Dropout is a form of missing data; the pattern alone does not establish MCAR, MAR, or MNAR.



## Wide-to-long transformation: weeks 5–24

```{r}
time.data2 <- reshape(time.data, direction = "long", 
                     varying = c("week_5","week_6","week_7","week_8","week_9",
                                 "week_10","week_11","week_12","week_13","week_14",
                                 "week_15","week_16","week_17","week_18","week_19",
                                 "week_20","week_21","week_22","week_23","week_24"),
                     idvar = c("pear_id","tree_id","ecotr_id","species","climate"),
                     timevar = "time", times = 5:24, v.names = "quality")
str(time.data2)
# write.csv(time.data2,"data/processed/longitudinal_data.csv",
#           row.names = FALSE)
```

## Dropout timing and retention by climate and variety

```{r}
time.data2 %>%
  group_by(species) %>%
  summarise(Total_NA = sum(is.na(quality)))

dropout_patterns <- time.data2 %>%
  filter(!is.na(quality)) %>%
  group_by(pear_id, species, climate) %>%
  summarise(last_obs_week = max(time), .groups = "drop")

baseline <- dropout_patterns %>%
  count(species, climate, name = "N")

risk_data <- dropout_patterns %>%
  crossing(Week = 5:24) %>% 
  group_by(species, climate, Week) %>%
  summarise(n = sum(last_obs_week >= Week), .groups = "drop") %>%
  left_join(baseline,  by = c("species", "climate")) %>%
  mutate(prop = n / N,
    percent = round(100 * prop))

table_14_1 <- risk_data %>%
  mutate(value = paste0(n, " (", percent, ")")) %>%
  select(Week, climate, species, value) %>%
  pivot_wider(names_from = species, values_from = value)

risk_data %>%
  group_by(species, Week) %>%
  summarise(n = sum(n),
            N = sum(N),
            percent = round(100 * n / N),
            .groups = "drop") %>%
  mutate(value = paste0(n, " (", percent, ")")) %>%
  select(Week, species, value) %>%
  pivot_wider(names_from = species, values_from = value)

ggplot(risk_data, aes(x = Week, y = prop, color = species)) +
  geom_step(size = 1) +
  facet_wrap(~ climate) +
  labs(x = "Weeks in study", y = "Proportion still in study") +
  theme_bw()
```

## Descriptive Kaplan–Meier retention curves

```{r}
library(survival)
surv_obj <- Surv(
  time = dropout_patterns$last_obs_week,
  event = dropout_patterns$last_obs_week < 24)
fit <- survfit(surv_obj ~ species, data = dropout_patterns)

plot(fit, col = 1:2, lty = 1:2,
     xlim = c(5,25), ylim = c(0.85, 1.0),
     xlab = "week", ylab = "proportion still in study")
legend("topright", legend = unique(dropout_patterns$species),
       col = 1:2, lty = 1:2)
```


## Size distributions and mean growth trajectories

```{r}
hist(time.data2$quality)
ggplot(time.data2) +
  geom_histogram(aes(quality),bins = 30, fill = "skyblue", colour = "black")+
  geom_vline(aes(xintercept = mean(time.data2$quality, na.rm=TRUE)), color = "red", 
             linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median(time.data2$quality, na.rm=TRUE)), color = "blue", 
             linetype = "dotted", size = 1) +
  labs(y="frequency") +
  scale_x_continuous(breaks = sort(c(pretty(time.data2$quality))))+
  theme_bw()

ggplot(time.data2, aes(x = time, y = quality, color = species, fill = species)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.3) +
  labs(y="average size of pears") +
  theme_bw()

ggplot(time.data2, aes(time,quality,colour = species))+
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(fill = species), 
               alpha = 0.2, colour = NA) +
  theme_bw()
```

## Individual pear trajectories

```{r}
my_cols <- c("salmon", "steelblue", "darkgrey", "darkgreen")
my_colors<-viridis(20)
my_lty <-c(5:24)
my_pch <-c(5:24)
```

## Growth by variety, ecotron and tree

```{r}
xyplot(quality ~ time | species,
       groups = pear_id,
       type = "l",
       col = my_colors,
       lty = my_lty,
        xlim = c(5, 24),
       data = time.data2)

xyplot(quality ~ time | climate,
       groups = pear_id,
       type = "l",
       xlim = c(5, 24),
       ylab = "size of pears",
       data = time.data2)

xyplot(quality ~ time | as.factor(ecotr_id),
       groups = pear_id,
       type = "l",
       col = my_colors,
       lty = my_lty,
        xlim = c(5, 24),
       data = time.data2)

xyplot(quality ~ time | as.factor(tree_id),
       groups = pear_id,
       type = "l",
       xlim = c(5, 24),
       ylab = "size of pears",
       data = time.data2)
```

## Tree–ecotron membership

```{r}
table(time.data2$ecotr_id,time.data2$tree_id)
```

## Weekly distributions and ecotron trajectories

```{r}
bwplot(quality ~ as.factor(time) | climate, 
       data = time.data2,
       xlab = "week",
       ylab = "size of pears (cm)",
       )

time.data2 %>% 
  group_by(time, ecotr_id, climate) %>% 
  summarise(mean_quality = mean(quality, na.rm = TRUE),
            sd_quality = sd(quality, na.rm = TRUE),  
            n_obs = sum(!is.na(quality)),
            se_quality = sd_quality / sqrt(n_obs),
            .groups = "drop") %>% 
  ggplot(aes(x = time, y = mean_quality, colour = climate, group = ecotr_id)) +
  geom_errorbar(aes(ymin = mean_quality - se_quality, 
                    ymax = mean_quality + se_quality))+
  geom_line() +
  geom_point()+
  theme_bw() +
  labs(x = "week",
       y = "average size of pears",
       colour = "climate")

ggplot(time.data2,aes(x = time, y = quality, colour = climate, group = pear_id)) +
  geom_line() +
  theme_bw() +
  labs(x = "week",
       y = "size of pears",
       colour = "climate")
```




## Soil source inspection

```{r}
soil.data <- read.csv("data/raw/G12.soil.data.csv",
                      sep = ",")
str(soil.data)
```

## Complete-case sensitivity dataset

```{r}
time.data3 <- na.omit(time.data)
sapply(time.data3, function(x) sum(is.na(x)))
```

## Complete-case long-format export

```{r}
time.data.long <- reshape(time.data3, direction = "long", 
                     varying = c("week_5","week_6","week_7","week_8","week_9",
                                 "week_10","week_11","week_12","week_13","week_14",
                                 "week_15","week_16","week_17","week_18","week_19",
                                 "week_20","week_21","week_22","week_23","week_24"),
                     idvar = c("pear_id","tree_id","ecotr_id","species","climate"),
                     timevar = "time", times = 5:24, v.names = "quality")
str(time.data.long)

# write.csv(time.data.long,"data/processed/time_data_clean.csv",
#           row.names = FALSE)
```






















